Proposal for Analyzing Language Models: Separatablity of Syntax and Semantics

### 2022-04-28 proposal/proposal.tex
fixing some citation

The boundary between semantics and syntax has always been a hotly debated topic in linguistics since generative syntax, but do NLP models make such a distinction? The objective of this project is to explore BERT\rq s reliance on certain syntactic information when handling a semantic task, and vice versa. To achieve our goal, we will construct a linear probing system for a task and then employ Iterative Nullspace Projection (INLP) \citep{inlp} to generate a new embedding devoid of information learned from probing task. Then we will measure the performance of another task on this new embedding.

The design of our probing procedure follows \citealp{amnesia}, who employed Iterative Nullspace Projection (INLP from here) to investigate whether BERT uses part-of-speech (PoS) information when solving language modeling (LM) tasks. A novel method for removing information from a embedding, INLP iteratively trains a linear model on a specific task, and projects the input on the nullspace of the linear model. Our objective is that, by applying the INLP procedure to a syntactic task, we are able to separate the representation into a syntactic space and a non-syntactic space. Then we can compare the performance of a semantic task in whole space and only in the non-syntactic space to see if BERT is using syntactic information when performing the semantic task. Conversely, we can also first probe a semantic task and then measure with a syntactic task.

To evaluate the separability of syntax and semantics, we will need two task, one task for the probing system and INLP procedure, one task for evaluating performance on embeddings before and after INLP. We choose Combinatory Categorical Grammar (CCG) tagging as the syntactic task and Semantic Role Labeling (SRL) as the semantic task. Section \ref{sec:method} provides description of the tasks and the construction of the experiment pipeline. Section \ref{sec:result} provides possible results and interpretations. Finally, section \ref{sec:management} provides project management details including timeline and division of labor.

@@

We construct two separate probing tasks to isolate the syntactic and semantic information in word-level BERT embeddings\citep{bert}. The embeddings are separated into syntactic and non-syntactic, and semantic and non-semantic components via Iterative Null-Space Projection (INLP), which is described in section 2.1. These embedding components are then combined to form new embeddings, which are evaluated on the same tasks that were used for probing. If time permits, we will also evaluate how these new embeddings perform on the language modeling task. 
%cite BERT paper maybe?
%cite INLP paper

>>>>

The boundary between semantics and syntax has always been a hotly debated topic in linguistics since generative syntax, but do NLP models make such a distinction? The objective of this project is to explore BERT\rq s\citep{bert}reliance on certain syntactic information when handling a semantic task, and vice versa. To achieve our goal, we will construct a linear probing system for a task and then employ Iterative Nullspace Projection (INLP from here on) \citep{inlp} to generate a new embedding devoid of information learned from probing task. Then we will measure the performance of another task on this new embedding.

The design of our probing procedure follows \citealp{amnesia}, who employed INLP to investigate whether BERT uses part-of-speech (PoS) information when solving language modeling (LM) tasks. A novel method for removing information from a embedding, INLP iteratively trains a linear model on a specific task, and projects the input on the nullspace of the linear model. Our objective is that, by applying the INLP procedure to a syntactic task, we are able to separate the representation into a syntactic space and a non-syntactic space. Then we can compare the performance of a semantic task in whole space and only in the non-syntactic space to see if BERT is using syntactic information when performing the semantic task. Conversely, we can also first probe a semantic task and then measure with a syntactic task.

To evaluate the separability of syntax and semantics, we will need two task, one task for the probing system and INLP procedure, one task for evaluating performance on embeddings before and after INLP. We choose Combinatory Categorical Grammar (CCG from here on) tagging \citep{ccg-bank} as the syntactic task and Semantic Role Labeling (SRL from here on) as the semantic task. Section \ref{sec:method} provides description of the tasks and the construction of the experiment pipeline. Section \ref{sec:result} provides possible results and interpretations. Finally, section \ref{sec:management} provides project management details including timeline and division of labor.

@@

We construct two separate probing tasks to isolate the syntactic and semantic information in word-level BERT embeddings. The embeddings are separated into syntactic and non-syntactic, and semantic and non-semantic components via INLP which is described in section 2.1. These embedding components are then combined to form new embeddings, which are evaluated on the same tasks that were used for probing. If time permits, we will also evaluate how these new embeddings perform on the language modeling task. 
%cite BERT paper maybe?
%cite INLP paper
